Efficient FPGA Implementation of\\ Digit Parallel Online Arithmetic Operators

### 2014-10-05 FPT2014/bare_conf.tex
add authors and shrink to 8 pages

% author names and affiliations
%\author{Authors removed for blind review\\ \\ \\}
%\author{\IEEEauthorblockN{Kan Shi, David Boland, and George A. Constantinides}
%\IEEEauthorblockA{Department of Electrical and Electronic Engineering\\
%Imperial College London\\London, UK\\
%\{k.shi11, david.boland03,g.constantinides\}@imperial.ac.uk}
%}

@@

\begin{abstract}
%\boldmath
Online arithmetic has been widely studied for ASIC implementation. Online components were originally designed to perform computations in digit serial with most significant digit (MSD) first, resulting in the ability to chain arithmetic operators together for low latency. More recently, research has shown that digit parallel online operators can fail more gracefully when operating beyond the deterministic clocking region in comparison to operators with conventional arithmetic. Unfortunately, the utilization of online arithmetic operators in the past has required a large area overhead for FPGA implementation. In this paper, we propose novel approaches to implement the key primitives of online arithmetic, adders and multipliers, efficiently on modern Xilinx FPGAs with 6-input LUTs and carry resources. We demonstrate experimentally that in comparison to a direct RTL synthesis, the proposed architectures achieve slice savings of over $67\%$ and $69\%$, and speed-ups of over $1.2\times$ and $1.5\times$ for adders and multipliers, respectively. As a result, the area overheads of using online adders and multipliers in place of traditional arithmetic primitives is reduced from $8.41\times$ and $8.11\times$ to $1.88\times$ and $1.84\times$ respectively. Finally, because an online multiplier generates MSDs first, we also demonstrate the method to create an online multiplier with a reduced precision output that is smaller than a traditional multiplier producing the same result. We show that this can lead to silicon area savings of up to $56\%$.
% Furthermore in comparison to standard arithmetic primitives, this translates to reduction of the area overheads from up to $8.41\times$ to $1.88\times$ for adders, and from $8.11\times$ to $1.84\times$ for multipliers using direct implementation, to $1.88\times$ and $0.54\times$ using our approaches.

@@

\section{Introduction}\label{Sec:Intro}
In the conventional form of computer arithmetic, the computation results are generated either from the least significant digit (LSD), e.g. addition and multiplication, or from the most significant digit (MSD), e.g. division and square root. This inconsistency in computing directions potentially results in large latency when propagating data between different operations. Online arithmetic was designed to solve this problem \cite{Ercegovac_OnlineOverview,Ercegovac_Book}. With online arithmetic, both inputs and outputs are processed in a MSD-first manner. This enables parallelism among multiple operations, and the overall computation latency can be significantly reduced. A brief overview of online arithmetic is given in Section~\ref{Sec:Background}.

Recent research has demonstrated another key advantage of online arithmetic: that it allows graceful degradation when operating beyond the deterministic clocking region,  \textit{i.e.} when overclocking \cite{SKDAC14_REVIEW}. This theme of research stems from the inspiration that for certain applications, releasing the absolute accuracy requirements could lead to significant performance improvements. However, traditional forms of computer arithmetic do not fail gracefully when timing violation happens, because timing errors tend to affect MSDs of the results. In comparison, online arithmetic is more ``overclocking friendly'', as timing errors only occur at LSDs of the results.

Despite its attractiveness, the use of online arithmetic on existing FPGAs is still limited due to a large area overhead. For standard arithmetic, because it is inefficient and slow to map arithmetic circuits solely onto look-up-tables (LUTs), both main commercial FPGA vendors have introduced dedicated carry logic~\cite{Virtex6}. Significant prior work explores methods that can effectively map circuits to the carry logic for performance improvements~\cite{FPL10_FPGA_CarryChain}, or proposes alternative carry logic for FPGAs~\cite{FPGA_CarryChain_New1,FPGA_CarryChain_New2}. Hard DSP blocks are also included in modern architectures to improve the area and performance of multiplications and multiply accumulates~\cite{XilinxDSP}. However, most of these approaches are designed to accelerate computations with the conventional form of arithmetic.

For online arithmetic, two major research directions have been investigated. Firstly, there has been some research into FPGA implementations of online adders; a brief review of relevant work in this area is presented in Section~\ref{Sec:OA_FPGA}. Secondly, there are existing approaches to build general purpose online adders specifically for FPGAs with 4-input LUTs and with two LUTs within a slice or a logic element (LE), such as the Xilinx Spartan~3 series \cite{XilinxSpartan} and the Altera Cyclone III \cite{AlteraCyclone}. However, as will be discussed in Section~\ref{Sec:OA_FPGA}, these approaches cannot be directly applied on many modern FPGAs, which typically have 6-input LUTs and four LUTs in a slice, such as the Xilinx Virtex series FPGAs and all Xilinx 7 series FPGAs~\cite{Virtex7}. Moreover, the move to the 6-LUT on the basic architecture opens new opportunities for further optimization.
%However, the implementations of other key arithmetic primitives, such as online multipliers, lack exploration.

In this paper, both issues will be addressed. In Section~\ref{Sec:OA_FPGA}, we propose a novel approach to map digit parallel online adders onto FPGAs with 6-LUTs. The mapping is based on the fast carry logic and ensures the available logic resources within a slice are fully utilized. In Section~\ref{Sec:OM_FPGA}, we optimize the online multiplication algorithm to yield an efficient digit parallel online multiplier for FPGAs. We demonstrate experimentally on Xilinx Virtex-6 FPGAs that the proposed designs achieve significant area reduction and performance gain over their original implementations. For online adders, our method achieves  slice savings over $67\%$ and  frequency speed-ups of over $1.2\times$ in comparison to a direct implementation. In comparison to the standard ripple-carry adder, the area overhead drops from up to $8.41\times$ to $1.88\times$ using the proposed online adder. For online multipliers, the slice utilization of the proposed architecture drops of over $69\%$, with a frequency speed-up over $1.5\times$ in comparison to a direct implementation. When compared to the multiplier with standard arithmetic, the direct implementation has an area overhead up to $8.11\times$, which drops to $1.84\times$ when using the proposed new architecture.

Our approach has the additional advantage of being able to create a correctly rounded multiplier with a reduced output precision using less hardware than a full multiplier, which multiplies two $N$-digit numbers and produces an output of $2N$ digits. However, in many cases the full precision product is not required. In such cases, a reduced precision output is created either by truncating or rounding the LSDs. Although there has been research into saving area by avoiding the creation of unnecessary LSDs~\cite{TheoTC}, these techniques still require extensive simulation to guarantee correctness. In contract, since online multipliers compute the output from the MSD first, hardware can be saved by not generating the LSDs in the first place. As an example, if we are only interested in the most significant half of the product, we can create a digit parallel online multiplier that saves up to $56\%$ slice usage in comparison to a standard multiplier with truncated output.

@@

In summary, the main contributions of this paper are:
\begin{enumerate}
    \item Efficient methods to map online adders and multipliers to modern FPGAs;
    \item A method to implement correctly-rounded online multipliers for a chosen output precision;
    \item Demonstration of the performance improvements of online operators in terms of area and frequency.
\end{enumerate}

@@

\section{Background: Online Arithmetic}\label{Sec:Background}
\subsection{Key Features of Online Arithmetic}

Online arithmetic has been widely used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.
%
\begin{figure}[htbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.pdf}
  \vspace{-4ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-1ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible only if a redundant number system is used. Normally there are two most commonly used redundant number representations: carry-save (CS) \cite{CSadder} and signed-digit (SD) \cite{RedundantNumber}. With SD representation, each digit is represented using a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. In comparison, the standard non-redundant representation only uses a digit set $\{0,\cdots,r-1\}$. Thus a standard number corresponds to several possible redundant representations. For example, the binary number $0.011$ can be represented in SD form as $0.1\overline{1}1$, $0.10\overline{1}$ or $0.011$ among many other possible representations.

Due to the redundancy, the MSDs of the result can be calculated using partial information from both inputs. Then the value of the number can be revised using the subsequent digits, because each number has multiple representations.

\subsection{Binary Online Addition}
Adders serve as a critical building block for arithmetic operations. To perform digit-parallel online addition, a redundant adder can be used directly. The structure of an online adder where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes three inputs and generates two outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labelled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only two FA delays for any operand word-length, with the cost of one extra FA for each digit of operands. This makes the online adder suitable for building up more complex arithmetic operators such as multipliers to accelerate the sum of partial products \cite{RedundantMult_1987}.
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder2.pdf}
    \vspace{-6ex}
	\caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
    \vspace{-2ex}
	\label{Fig:Radix2SD_adder}
\end{figure}

@@

\subsection{Binary Online Multiplication}
Multiplication is another key arithmetic operator. Typically, online multiplication is performed in a recursive digit serial manner, as illustrated in Algorithm~\ref{Algorithm:OnlineMult} \cite{Ercegovac_OnlineMult} where both inputs and outputs are $N$-digit numbers as given in~(\ref{Eq:Online_Operands}). For a given iteration $j$, the product digit $z_j$ is generated MSD-first through a selection function $sel()$. For any radix $r$ and chosen digit set, there exits an appropriate selection method and a value of $\delta$ which ensure convergence. As the binary radix is used most commonly in computer arithmetic, we keep $r=2$ throughout this paper with the corresponding redundant digit set $\{\overline{1},0,1\}$. In this case $sel()$ is given by (\ref{Eq:SelFunc_OM}), and the selection is based on two integer digits and one fractional digit of $W_{[j]}$ \cite{Ercegovac_OnlineMult}.
%
\begin{eqnarray}\label{Eq:SelFunc_OM}
\small
  sel(W_{[j]})=\begin{cases}
    1 & \text{ if } W_{[j]} \geqslant \frac{1}{2} \\
    0 & \text{ if } -\frac{1}{2}\leqslant W_{[j]}<\frac{1}{2} \\
    \overline{1} & \text{ if } W_{[j]}<-\frac{1}{2}
  \end{cases}
\normalsize
\end{eqnarray}

\begin{algorithm}[tbp]
  \caption{Online Multiplication}  \label{Algorithm:OnlineMult}
  \begin{algorithmic}[1]
   \State \textbf{Initialization:}~$X_{[-\delta]}=Y_{[-\delta]}=P_{[-\delta]}=0$\vspace{.5ex}
    \For{$j=-\delta,~-\delta+1,~\cdots,~2N-1$}\vspace{.5ex}
        \State $H_{[j]} \leftarrow r^{-\delta}\left(x_{j+\delta+1}\cdot Y_{[j+1]}+y_{j+\delta+1}\cdot X_{[j]}\right)$\vspace{.5ex}
        \State $W_{[j]} \leftarrow P_{[j]} + H_{[j]}$\vspace{.5ex}
        \State $z_{j}~ \leftarrow sel(W_{[j]})$\vspace{.5ex}
        \State $P_{[j+1]} \leftarrow r\left(W_{[j]}-Z_{[j]}\right)$\vspace{.5ex}
    \EndFor
  \end{algorithmic}
\end{algorithm}
\vspace{-1ex}

It is worth noting that the precision of the product $Z$ can be determined by directly truncating from the LSDs, since there is no carry propagation from the LSD to the MSD. This is useful especially when the multiplication product is used by successive computations, where often only the first $N$ digits of the results are used. In comparison, the product is generated starting from the LSD with conventional arithmetic. Hence the successive operations cannot commence until the result with full-precision is generated. This feature of online arithmetic has inspired research about dynamically control of computation precision in numerous applications~\cite{Online_Truncation}.

@@

In Section~\ref{Sec:OM_FPGA}, we will first describe an optimized online multiplication algorithm and its FPGA implementation, which are designed specifically targeting digit parallel operations. Then we provide a modified structure which only generates the most significant half of the product in order to achieve extra area savings.

@@

\section{Digit Parallel Online Adder for FPGAs}\label{Sec:OA_FPGA}
\subsection{Related Works}
There have been previous works about FPGA implementation of digit parallel online adders. From the literature, the existing approaches can be classified into three types:

@@

Type 2 examines the use of counters to sum $k$ $N$-digit numbers where $k>2$. Instead, we focus on two-operand online addition that performs $Z=X+Y$, as shown in Fig.~\ref{Fig:Radix2SD_adder}. In comparison to type 3, this paper attempts to obtain the best performance from mainstream FPGAs instead of suggesting ways to modify the FPGA fabric.

Specifically in type 1, both works took advantage of the built-in carry resources in FPGAs. Conventionally the ASIC implementation of online adders is based on 4:2 compressors, as outlined in Fig.~\ref{Fig:PreviousWork} within the gray background. However, directly applying this approach in the FPGAs could be less efficient. This is because there is no carry propagation between the two FAs within a 4:2 compressor, and the net delay between them can be large. Instead, Kamp et al \cite{FPT09RA} and Hormigo et al \cite{ASAP09RA} described very similar mapping techniques for online adders with SD representations and CS representations, respectively. They share the common idea of building the online adder based on the logic block (\texttt{LB}) as highlighted within the dotted circle in Fig.~\ref{Fig:PreviousWork}. In this case, the fast-carry logic in the FPGA can be employed, and the delay between the two 3:2 compressors can be greatly reduced.
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder_42comp_Spartan2.pdf}
	\vspace{-3ex}
	\caption{Map the online adder onto Spartan FPGAs using the fast-carry resources. The gray background highlights the 4:2 compressor. Dotted circle indicates the logic block (LB) which can be mapped to the FPGA using the carry resources.}
	\vspace{-4ex}
	\label{Fig:PreviousWork}
\end{figure}

However, we notice the major limitations of both approaches that they only target FPGAs with 4-input LUTs (4-LUT) and two LUTs within a logic slice, such as the Xilinx Spartan series and the Altera Cyclone series. This is naturally reasonable because one \texttt{LB} can be mapped to a single slice. Nevertheless, for FPGAs with 6-input LUTs (6-LUT) and four LUTs in a slice, such as the Xilinx Virtex series and all Xilinx 7 series FPGAs, directly applying these approaches will result in either resource waste or wrong results. For instance, if two \texttt{LB}s are mapped to a slice with four LUTs as seen in Fig.~\ref{Fig:PreWorkVirtex6}, the outputs of \texttt{LB1} should not be used because the its carry input cannot be explicitly initialized. Due to the same reason, the output of \texttt{XORCY2} should not be used.
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.45\textwidth]{./Figures/SDadder_FastCarry_PreviousWork2.pdf}
	\vspace{-2ex}
	\caption{An example illustrating that direct applying approaches in \cite{FPT09RA} on a Virtex-6 FPGA will results in faulty outputs, because the carry input of \texttt{LB1} cannot be explicitly initialized.}
	\label{Fig:PreWorkVirtex6}
    \vspace{-2ex}
\end{figure}

\subsection{Proposed Mapping Method}
To tackle this problem, we first present an alternative structure of the online adder to enable an efficient FPGA mapping. The structure of a 4-digit online adder is given as an example in Fig.~\ref{Fig:ModifiedOA}. In this equivalent structure, the first FA in each 4:2 compressor is split into two parts, which only generate carry and sum respectively. In this case they can be mapped individually on two LUTs. The second FA, which generates the outputs, is unchanged and can be implemented using the fast carry logic.

>>>>

% author names and affiliations
%\author{Authors removed for blind review\\ \\ \\}
\author{
\IEEEauthorblockN{Kan Shi}
\IEEEauthorblockA{Department of Electrical and\\ Electronic Engineering\\
Imperial College London\\London, SW7 2AZ, UK\\
k.shi11@imperial.ac.uk}
\and
\IEEEauthorblockN{David Boland}
\IEEEauthorblockA{Department of Electrical and\\ Computer Systems Engineering\\
Monash University\\Clayton, VIC 3800, Australia\\
david.boland@monash.edu
}
\and
\IEEEauthorblockN{George A. Constantinides}
\IEEEauthorblockA{Department of Electrical and\\ Electronic Engineering\\
Imperial College London\\London, SW7 2AZ, UK\\
g.constantinides@imperial.ac.uk}
%\{k.shi11, david.boland03,g.constantinides\}@imperial.ac.uk}
}

@@

\begin{abstract}
%\boldmath
Online arithmetic has been widely studied for ASIC implementation. Online components were originally designed to perform computations in digit serial with most significant digit (MSD) first, resulting in the ability to chain arithmetic operators together for low latency. More recently, research has shown that digit parallel online operators can fail more gracefully when operating beyond the deterministic clocking region in comparison to operators with conventional arithmetic. Unfortunately, the utilization of online arithmetic operators in the past has required a large area overhead for FPGA implementation. In this paper, we propose novel approaches to implement the key primitives of online arithmetic, adders and multipliers, efficiently on modern Xilinx FPGAs with 6-input LUTs and carry resources. We demonstrate experimentally that in comparison to a direct RTL synthesis, the proposed architectures achieve slice savings of over $67\%$ and $69\%$, and speed-ups of over $1.2\times$ and $1.5\times$ for adders and multipliers, respectively. As a result, the area overheads of using online adders and multipliers in place of traditional arithmetic primitives is reduced from $8.41\times$ and $8.11\times$ to $1.88\times$ and $1.84\times$ respectively. Finally, because an online multiplier generates MSDs first, we also demonstrate the method to create an online multiplier with a reduced precision output that is smaller than a traditional multiplier producing the same result. We show that this can lead to silicon area savings of up to $56\%$.\vspace{-0.5ex}
% Furthermore in comparison to standard arithmetic primitives, this translates to reduction of the area overheads from up to $8.41\times$ to $1.88\times$ for adders, and from $8.11\times$ to $1.84\times$ for multipliers using direct implementation, to $1.88\times$ and $0.54\times$ using our approaches.

@@

\section{Introduction}\label{Sec:Intro}\vspace{-0.5ex}
In the conventional form of computer arithmetic, the computation results are generated either from the least significant digit (LSD), e.g. addition and multiplication, or from the most significant digit (MSD), e.g. division and square root. This inconsistency in computing directions potentially results in large latency when propagating data between different operations. Online arithmetic was designed to solve this problem \cite{Ercegovac_OnlineOverview,Ercegovac_Book}. With online arithmetic, both inputs and outputs are processed in a MSD-first manner. This enables parallelism among multiple operations, and the overall computation latency can be significantly reduced. A brief overview of online arithmetic is given in Section~\ref{Sec:Background}.\vspace{-0.5ex}

Recent research has demonstrated another key advantage of online arithmetic: that it allows graceful degradation when operating beyond the deterministic clocking region,  \textit{i.e.} when overclocking \cite{SKDAC14}. This theme of research stems from the inspiration that for certain applications, releasing the absolute accuracy requirements could lead to significant performance improvements. However, traditional forms of computer arithmetic do not fail gracefully when timing violation happens, because timing errors tend to affect MSDs of the results. In comparison, online arithmetic is more ``overclocking friendly'', as timing errors only occur at LSDs of the results.\vspace{-0.5ex}

Despite its attractiveness, the use of online arithmetic on existing FPGAs is still limited due to a large area overhead. For standard arithmetic, because it is inefficient and slow to map arithmetic circuits solely onto look-up-tables (LUTs), both main commercial FPGA vendors have introduced dedicated carry logic~\cite{Virtex6}. Significant prior work explores methods that can effectively map circuits to the carry logic for performance improvements~\cite{FPL10_FPGA_CarryChain}, or proposes alternative carry logic for FPGAs~\cite{FPGA_CarryChain_New1,FPGA_CarryChain_New2}. Hard DSP blocks are also included in modern architectures to improve the area and performance of multiplications and multiply accumulates. However, most of these approaches are designed to accelerate computations with the conventional form of arithmetic.\vspace{-0.5ex}

For online arithmetic, two major research directions have been investigated. Firstly, there has been some research into FPGA implementations of online adders; a brief review of relevant work in this area is presented in Section~\ref{Sec:OA_FPGA}. Secondly, there are existing approaches to build general purpose online adders specifically for FPGAs with 4-input LUTs and with two LUTs within a slice or a logic element (LE), such as the Xilinx Spartan~3 series \cite{XilinxSpartan} and the Altera Cyclone III \cite{AlteraCyclone}. However, as will be discussed in Section~\ref{Sec:OA_FPGA}, these approaches cannot be directly applied on many modern FPGAs, which typically have 6-input LUTs and four LUTs in a slice, such as the Xilinx Virtex series FPGAs and all Xilinx 7 series FPGAs~\cite{Virtex7}. Moreover, the move to the 6-LUT on the basic architecture opens new opportunities for further optimization.\vspace{-0.5ex}
%However, the implementations of other key arithmetic primitives, such as online multipliers, lack exploration.

In this paper, both issues will be addressed. In Section~\ref{Sec:OA_FPGA}, we propose a novel approach to map digit parallel online adders onto FPGAs with 6-LUTs. The mapping is based on the fast carry logic and ensures the available logic resources within a slice are fully utilized. In Section~\ref{Sec:OM_FPGA}, we optimize the online multiplication algorithm to yield an efficient digit parallel online multiplier for FPGAs. We demonstrate experimentally on Xilinx Virtex-6 FPGAs that the proposed designs achieve significant area reduction and performance gain over their original implementations. For online adders, our method achieves  slice savings over $67\%$ and  frequency speed-ups of over $1.2\times$ in comparison to a direct implementation. In comparison to the standard ripple-carry adder, the area overhead drops from up to $8.41\times$ to $1.88\times$ using the proposed online adder. For online multipliers, the slice utilization of the proposed architecture drops of over $69\%$, with a frequency speed-up over $1.5\times$ in comparison to a direct implementation. When compared to the multiplier with standard arithmetic, the direct implementation has an area overhead up to $8.11\times$, which drops to $1.84\times$ when using the proposed new architecture.\vspace{-0.5ex}

Our approach has the additional advantage of being able to create a correctly rounded multiplier with a reduced output precision using less hardware than a full multiplier, which multiplies two $N$-digit numbers and produces an output of $2N$ digits. However, in many cases the full precision product is not required. In such cases, a reduced precision output is created either by truncating or rounding the LSDs. Although there has been research into saving area by avoiding the creation of unnecessary LSDs~\cite{TheoTC}, these techniques still require extensive simulation to guarantee correctness. In contract, since online multipliers compute the output from the MSD first, hardware can be saved by not generating the LSDs in the first place. As an example, if we are only interested in the most significant half of the product, we can create a digit parallel online multiplier that saves up to $56\%$ slice usage in comparison to a standard multiplier with truncated output.\vspace{-0.5ex}

@@

In summary, the main contributions of this paper are:\vspace{-0.5ex}
\begin{enumerate}
    \item Efficient methods to map online adders and multipliers to modern FPGAs;
    \item A method to implement correctly-rounded online multipliers for a chosen output precision;
    \item Demonstration of the performance improvements of online operators in terms of area and frequency.
\end{enumerate}\vspace{-1ex}

@@

\section{Background: Online Arithmetic}\label{Sec:Background}\vspace{-0.5ex}
\subsection{Key Features of Online Arithmetic}\vspace{-0.5ex}

Online arithmetic has been widely used in numerous applications such as signal processing and control algorithms~\cite{Online_FPGADSP,Online_Control}. Online arithmetic was originally designed for digit-serial operation, as illustrated in Fig~\ref{Fig:OnlineDataFlow}. It can be seen that in order to generate the first output digit, $\delta$ digits of inputs are required, where $\delta$ is called the ``online delay''. $\delta$ is normally a small constant, which is independent of the precision. For ease of discussion, for the rest of this paper, the input data is assumed to be fixed point numbers in the range $(-1,1)$. Based on this premise, the online representation of $N$-digit operands and result at iteration $j$ are given by (\ref{Eq:Online_Operands}), where $j\in[-\delta,N-1]$ and $r$ denotes the radix \cite{Ercegovac_Book}.\vspace{-0.5ex}
%
\begin{figure}[tbp]
  \centering
  %\vspace{-2.5ex}
  \includegraphics[width=.48\textwidth]{./Figures/OnlineArithmetic_DataFlow.pdf}
  \vspace{-4ex}
  \caption{Dataflow in digit-serial online arithmetic, in which both inputs and outputs are processed from the MSD to the LSD. $\delta$ denotes the online delay.}
  \vspace{-1ex}
  \label{Fig:OnlineDataFlow}
\end{figure}
%
\begin{eqnarray}\label{Eq:Online_Operands}
\footnotesize
  X_{[j]}=\sum_{i=1}^{j+\delta}x_ir^{-i},~Y_{[j]}=\sum_{i=1}^{j+\delta}y_ir^{-i},~Z_{[j]}=\sum_{i=1}^{j}z_ir^{-i}
\normalsize
\end{eqnarray}

MSD-first operation is possible only if a redundant number system is used. Normally there are two most commonly used redundant number representations: carry-save (CS) \cite{CSadder} and signed-digit (SD) \cite{RedundantNumber}. With SD representation, each digit is represented using a redundant digit set $\{-a, \cdots,-1,0, 1, \cdots, a\}$, where $a\in[r/2,r-1]$. In comparison, the standard non-redundant representation only uses a digit set $\{0,\cdots,r-1\}$. Thus a standard number corresponds to several possible redundant representations. For example, the binary number $0.011$ can be represented in SD form as $0.1\overline{1}1$, $0.10\overline{1}$ or $0.011$ among many other possible representations.\vspace{-0.5ex}
%
\vspace{-.5ex}
\begin{figure}[tbp]
  \centering
  \includegraphics[width=.5\textwidth]{./Figures/SDAdder2.pdf}
    \vspace{-7ex}
  \caption{An $N$-digit binary digit-parallel online adder. Both inputs and outputs are represented using SD representation. ``3:2'' denotes a 3:2 compressor.}
    \vspace{-2ex}
  \label{Fig:Radix2SD_adder}
\end{figure}

Due to the redundancy, the MSDs of the result can be calculated using partial information from both inputs. Then the value of the number can be revised using the subsequent digits, because each number has multiple representations.\vspace{-0.5ex}

\subsection{Binary Online Addition}\vspace{-0.5ex}
Adders serve as a critical building block for arithmetic operations. To perform digit-parallel online addition, a redundant adder can be used directly. The structure of an online adder where all signals represented with SD numbers of digit set $\{-1,0,1\}$ is shown in Fig.~\ref{Fig:Radix2SD_adder}. The module ``3:2'' denotes a 3:2 compressor, which takes three inputs and generates two outputs, and is logically equivalent to a full adder (FA). A major advantage of the redundant number system over the standard ripple-carry based arithmetic is that the propagation of carry is eliminated, resulting in a precision-independent computation time for addition. As labelled in Fig.~\ref{Fig:Radix2SD_adder}, ideally the computation delay of this adder is only two FA delays for any operand word-length, with the cost of one extra FA for each digit of operands. This makes the online adder suitable for building up more complex arithmetic operators such as multipliers to accelerate the sum of partial products \cite{RedundantMult_1987}.\vspace{-0.5ex}
%

@@

\subsection{Binary Online Multiplication}\vspace{-0.5ex}
Multiplication is another key arithmetic operator. Typically, online multiplication is performed in a recursive digit serial manner, as illustrated in Algorithm~\ref{Algorithm:OnlineMult} \cite{Ercegovac_OnlineMult} where both inputs and outputs are $N$-digit numbers as given in~(\ref{Eq:Online_Operands}). For a given iteration $j$, the product digit $z_j$ is generated MSD-first through a selection function $sel()$. For any radix $r$ and chosen digit set, there exits an appropriate selection method and a value of $\delta$ which ensure convergence. As the binary radix is used most commonly in computer arithmetic, we keep $r=2$ throughout this paper with the corresponding redundant digit set $\{\overline{1},0,1\}$. In this case $sel()$ is given by (\ref{Eq:SelFunc_OM}), and the selection is based on two integer digits and one fractional digit of $W_{[j]}$ \cite{Ercegovac_OnlineMult}.\vspace{-0.5ex}
%
\begin{eqnarray}\label{Eq:SelFunc_OM}
\small
  sel(W_{[j]})=\begin{cases}
    1 & \text{ if } W_{[j]} \geqslant \frac{1}{2} \\
    0 & \text{ if } -\frac{1}{2}\leqslant W_{[j]}<\frac{1}{2} \\
    \overline{1} & \text{ if } W_{[j]}<-\frac{1}{2}
  \end{cases}
\normalsize
\end{eqnarray}

\begin{algorithm}[tbp]
  \caption{Online Multiplication}  \label{Algorithm:OnlineMult}
  \begin{algorithmic}[1]
   \State \textbf{Initialization:}~$X_{[-\delta]}=Y_{[-\delta]}=P_{[-\delta]}=0$\vspace{.5ex}
    \For{$j=-\delta,~-\delta+1,~\cdots,~2N-1$}\vspace{.5ex}
        \State $H_{[j]} \leftarrow r^{-\delta}\left(x_{j+\delta+1}\cdot Y_{[j+1]}+y_{j+\delta+1}\cdot X_{[j]}\right)$\vspace{.5ex}
        \State $W_{[j]} \leftarrow P_{[j]} + H_{[j]}$\vspace{.5ex}
        \State $z_{j}~ \leftarrow sel(W_{[j]})$\vspace{.5ex}
        \State $P_{[j+1]} \leftarrow r\left(W_{[j]}-Z_{[j]}\right)$\vspace{.5ex}
    \EndFor
  \end{algorithmic}
\end{algorithm}
\vspace{-1.5ex}

It is worth noting that the precision of the product $Z$ can be determined by directly truncating from the LSDs, since there is no carry propagation from the LSD to the MSD. This is useful especially when the multiplication product is used by successive computations, where often only the first $N$ digits of the results are used. In comparison, the product is generated starting from the LSD with conventional arithmetic. Hence the successive operations cannot commence until the result with full-precision is generated. This feature of online arithmetic has inspired research about dynamically control of computation precision in numerous applications~\cite{Online_Truncation}.\vspace{-0.5ex}

@@

In Section~\ref{Sec:OM_FPGA}, we will first describe an optimized online multiplication algorithm and its FPGA implementation, which are designed specifically targeting digit parallel operations. Then we provide a modified structure which only generates the most significant half of the product in order to achieve extra area savings.\vspace{-0.5ex}

@@

\section{Digit Parallel Online Adder for FPGAs}\label{Sec:OA_FPGA}
\subsection{Related Works}\vspace{-0.5ex}
There have been previous works about FPGA implementation of digit parallel online adders. From the literature, the existing approaches can be classified into three types:\vspace{-0.5ex}

@@

Type 2 examines the use of counters to sum $k$ $N$-digit numbers where $k>2$. Instead, we focus on two-operand online addition that performs $Z=X+Y$, as shown in Fig.~\ref{Fig:Radix2SD_adder}. In comparison to type 3, this paper attempts to obtain the best performance from mainstream FPGAs instead of suggesting ways to modify the FPGA fabric.\vspace{-0.5ex}

Specifically in type 1, both works took advantage of the built-in carry resources in FPGAs. Conventionally the ASIC implementation of online adders is based on 4:2 compressors, as outlined in Fig.~\ref{Fig:PreviousWork} within the gray background. However, directly applying this approach in the FPGAs could be less efficient. This is because there is no carry propagation between the two FAs within a 4:2 compressor, and the net delay between them can be large. Instead, Kamp et al \cite{FPT09RA} and Hormigo et al \cite{ASAP09RA} described very similar mapping techniques for online adders with SD representations and CS representations, respectively. They share the common idea of building the online adder based on the logic block (\texttt{LB}) as highlighted within the dotted circle in Fig.~\ref{Fig:PreviousWork}. In this case, the fast-carry logic in the FPGA can be employed, and the delay between the two 3:2 compressors can be greatly reduced.
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.5\textwidth]{./Figures/SDAdder_42comp_Spartan2.pdf}
	\vspace{-4ex}
	\caption{Map the online adder onto Spartan FPGAs using the fast-carry resources. The gray background highlights the 4:2 compressor. Dotted circle indicates the logic block (LB) which can be mapped to the FPGA using the carry resources.}
	\vspace{-4ex}
	\label{Fig:PreviousWork}
\end{figure}

However, we notice the major limitations of both approaches that they only target FPGAs with 4-input LUTs (4-LUT) and two LUTs within a logic slice, such as the Xilinx Spartan series and the Altera Cyclone series. This is naturally reasonable because one \texttt{LB} can be mapped to a single slice. Nevertheless, for FPGAs with 6-input LUTs (6-LUT) and four LUTs in a slice, such as the Xilinx Virtex series and all Xilinx 7 series FPGAs, directly applying these approaches will result in either resource waste or wrong results. For instance, if two \texttt{LB}s are mapped to a slice with four LUTs as seen in Fig.~\ref{Fig:PreWorkVirtex6}, the outputs of \texttt{LB1} should not be used because its carry input cannot be explicitly initialized. Due to the same reason, the output of \texttt{XORCY2} should not be used.\vspace{-0.5ex}
%
\begin{figure}[tbp]
	\centering
	\includegraphics[width=.45\textwidth]{./Figures/SDadder_FastCarry_PreviousWork2.pdf}
	\vspace{-3ex}
	\caption{An example illustrating that direct applying approaches in \cite{FPT09RA} on a Virtex-6 FPGA will results in faulty outputs, because the carry input of \texttt{LB1} cannot be explicitly initialized.}
	\label{Fig:PreWorkVirtex6}
    \vspace{-2.5ex}
\end{figure}

\subsection{Proposed Mapping Method}\vspace{-0.5ex}
To tackle this problem, we first present an alternative structure of the online adder to enable an efficient FPGA mapping. The structure of a 4-digit online adder is given as an example in Fig.~\ref{Fig:ModifiedOA}. In this equivalent structure, the first FA in each 4:2 compressor is split into two parts, which only generate carry and sum respectively. In this case they can be mapped individually on two LUTs. The second FA, which generates the outputs, is unchanged and can be implemented using the fast carry logic.\vspace{-0.5ex}
